---
title: "ST558 Project 1"
author: "Aliyah Sheth"
format: html
editor: visual
---

## Introduction

This project develops functions for retrieving, processing, summarizing, and visualizing data from the Census Public Use Microdata Sample (PUMS) API. Data from the 2021 through 2024 surveys will be used. The project first processes Census variable metadata, then develops functions to query and combine PUMS data, followed by custom methods for weighted summaries and graphical displays.

## Part 1: Data Processing

### Parsing Variable Values

The first step is to obtain the variable descriptions for the Census PUMS data. The variable metadata are available as JSON files for each survey year. These files contain information about the variables and, where applicable, the values and labels associated with categorical variables.

The code below begins by querying the 2024 variable metadata to examine the structure of the returned JSON data. A status code of 200 indicates that the request was successful. 

```{r}
#| message: false

# Load packages used for this document
library(httr)
library(lubridate)

# URL for the 2024 PUMS variable metadata
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

# Query the Census API
response_2024 <- GET(url_2024)

# Check whether the request was successful
status_code(response_2024)
```

The returned JSON content is then parsed and examined to determine how the variable information is stored.

```{r}
# Parse the JSON response
metadata_2024 <- content(response_2024)

# Examine the top-level structure
names(metadata_2024)

# Examine the metadata for the SEX variable
metadata_2024$variables$SEX

# Examine the components stored for SEX
names(metadata_2024$variables$SEX)

# Variables required for the project
variables_2024 <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
  "STATE", "REGION", "DIVISION"
)

# Check whether any required variables are missing
variables_2024[
  !variables_2024 %in% names(metadata_2024$variables)
]
```
All variables required for the 2024 portion of the project are present in the Census metadata. 

#### Creating a Function to Parse Variable Values

After examining the structure of the 2024 variable metadata, a function is created to automate this process for the 2021 through 2024 survey years. The function identifies the appropriate state variable for each year, retrieves the corresponding JSON file, and extracts the value information for the variables required for this project.

```{r}
# Function to retrieve variable values for a specified survey year
get_variable_values <- function(year) {

  # Check that a valid survey year was provided
  if (!year %in% c(2021, 2022, 2023, 2024)) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }

  # Determine the appropriate state variable name
  if (year %in% c(2021, 2022)) {
    state_variable <- "ST"
  } else {
    state_variable <- "STATE"
  }

  # Define the variables needed for the project
  variables <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_variable, "REGION", "DIVISION"
  )

  # Create the URL for the selected year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )

  # Query and parse the variable metadata
  response <- GET(url)

  if (status_code(response) != 200) {
    stop("The Census API request was not successful.")
  }

  metadata <- content(response)

  # Create an empty list for the variable values
  variable_values <- list()
  
  # Extract the available values for each required variable
  for (variable in variables) {
   variable_values[variable] <- list(
     metadata$variables[[variable]]$values$item
   )
  }

  return(variable_values)
}
```

The function is tested using the 2024 survey year, to ensure it produces the expected results. 

```{r}
# Test the function using the 2024 survey year
variable_values_2024 <- get_variable_values(2024)

# Examine the variables contained in the resulting list
names(variable_values_2024)

# Examine the values returned for SEX
variable_values_2024$SEX
```

#### Applying the Function to Parse Variable Values

The function is then applied to each of the four survey years to create a separate list of variable values for 2021, 2022, 2023, and 2024.

```{r}
# Create variable value lists for each survey year
variable_values_2021 <- get_variable_values(2021)
variable_values_2022 <- get_variable_values(2022)
variable_values_2023 <- get_variable_values(2023)
variable_values_2024 <- get_variable_values(2024)

# Combine the four yearly lists
variable_values <- list(
  "2021" = variable_values_2021,
  "2022" = variable_values_2022,
  "2023" = variable_values_2023,
  "2024" = variable_values_2024
)

# Examine the survey years stored in the list
names(variable_values)

# Check the state variable used for each survey year
names(variable_values$`2021`)
names(variable_values$`2024`)
```

The results confirm that the function correctly uses "ST" for the 2021 and 2022 surveys and "STATE" for the 2023 and 2024 surveys.

```{r}
# Save the variable value lists for later use
saveRDS(
  variable_values,
  file = "data/census_variable_values.rds"
)

# Read the saved file to verify that it was created successfully
variable_values_check <- readRDS(
  "data/census_variable_values.rds"
)

names(variable_values_check)
```
The saved object contains the variable-value information for each survey year and can be loaded later when categorical and geographic variables are converted to their appropriate labels.

### Parsing API Data

The next step is to query the PUMS API for person-level data. Before creating a general function, a single API request is examined to understand the structure of the data returned by the Census API.

```{r}
# Retrieve Census API key from the environment
census_key <- Sys.getenv("CENSUS_API_KEY")

# Check that an API key is available
if (census_key == "") {
  stop("Census API key not found.")
}

# Create a sample 2024 PUMS API request
pums_url <- paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  "?get=SEX,AGEP,PWGTP",
  "&for=state:37",
  "&key=", census_key
)

# Query the Census API
pums_response <- GET(pums_url)

# Check whether the request was successful
status_code(pums_response)
```

A status code of 200 indicates that the PUMS API request was successful. 

The returned content is then parsed to examine how the observations are organized. The current Census Microdata API requires an API key for data queries. Therefore, an API key stored as an environment variable is used for the PUMS queries in this project.

```{r}
# Parse the API response
pums_content <- content(
  pums_response,
  as = "parsed",
  encoding = "UTF-8"
)

# Examine the first few elements of the returned object
pums_content[1:3]

# Examine the first row returned by the API
pums_content[[1]]
```

#### Creating a Helper Function to Parse PUMS Data

The PUMS API returns the column names as the first element of the parsed response and the observations as the remaining elements. A helper function is created to convert this structure into a tibble with the appropriate column names.

```{r}
# Helper function to convert parsed PUMS content into a tibble
parse_pums_response <- function(pums_content) {

  # Extract the column names from the first element
  column_names <- unlist(pums_content[[1]])

  # Create an empty matrix for the returned observations
  data_matrix <- matrix(
    NA_character_,
    nrow = length(pums_content) - 1,
    ncol = length(column_names)
  )

  # Add each observation to the matrix
  for (i in 2:length(pums_content)) {
    data_matrix[i - 1, ] <- unlist(
      pums_content[[i]]
    )
  }
  
  # Convert to a data frame
  pums_data <- as.data.frame(
    data_matrix,
    stringsAsFactors = FALSE
  )

  # Assign the column names
  names(pums_data) <- column_names

  # Convert to a tibble
  pums_data <- tibble::as_tibble(pums_data)

  return(pums_data)
}

# Test the helper function
pums_tibble <- parse_pums_response(pums_content)

# Examine the resulting tibble
head(pums_tibble)
```
The helper function successfully converts the parsed API response into a tibble. At this stage, all returned variables are stored as character values; appropriate numeric, time, and categorical conversions will be performed in the subsequent data-processing function.

#### Creating a Helper Function for Time Variables

The JWAP and JWDP variables represent times using coded ranges rather than exact times. To convert these variables into usable time values, a helper function is created to parse each time range and return its midpoint. A fixed reference date is used so that the resulting values can be stored as date-time objects; only the time component is relevant for this project.

```{r}
# Helper function to return the midpoint of a Census time range
parse_time_midpoint <- function(time_ranges) {

  # Create an empty time vector for the results
  midpoint_times <- rep(
    as.POSIXct(NA, tz = "UTC"),
    length(time_ranges)
  )

  # Process each time range
  for (i in seq_along(time_ranges)) {

    # Leave missing time values as NA
    if (is.na(time_ranges[i]) || time_ranges[i] == "") {
      next
    }

    # Remove periods from a.m. and p.m.
    clean_range <- gsub(
      "\\.",
      "",
      time_ranges[i]
    )

    # Separate the beginning and ending times
    time_parts <- strsplit(
      clean_range,
      " to "
    )[[1]]

    # Leave unexpected values as NA
    if (length(time_parts) != 2) {
      next
    }

    # Parse the beginning of the time interval
    start_time <- parse_date_time(
      paste("2000-01-01", time_parts[1]),
      orders = "Ymd IMp",
      tz = "UTC"
    )

    # Parse the end of the time interval
    end_time <- parse_date_time(
      paste("2000-01-01", time_parts[2]),
      orders = "Ymd IMp",
      tz = "UTC"
    )

    # Calculate the midpoint of the interval
    midpoint_times[i] <- start_time +
      (end_time - start_time) / 2
  }

  return(midpoint_times)
}

# Test the time midpoint helper function
time_test <- c(
  "10:20 a.m. to 10:24 a.m.",
  "4:10 a.m. to 4:19 a.m.",
  NA
)

parse_time_midpoint(time_test)
```

#### Creating the Single-Year PUMS Query Function

The next step is to create a function that allows the user to specify the survey year, numeric and categorical variables, and geographic level to return. The function checks that the requested inputs meet the project requirements before building and submitting the API request. The user can select State, Division, or Region and can request either a specific geography or all values within that geography.

```{r}
# Function to query one year of PUMS data
get_pums_data <- function(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = NULL
) {

  # Define valid survey years
  valid_years <- c(2021, 2022, 2023, 2024)

  # Check that exactly one valid year was provided
  if (length(year) != 1 || !year %in% valid_years) {
    stop("Year must be one of 2021, 2022, 2023, or 2024.")
  }

  # Define valid numeric variables
  valid_numeric_vars <- c(
    "AGEP", "GASP", "GRPIP",
    "JWAP", "JWDP", "JWMNP",
    "PWGTP"
  )

  # Check that all numeric variables are valid
  if (any(!numeric_vars %in% valid_numeric_vars)) {
    stop("One or more numeric variables are not valid.")
  }

  # Check that PWGTP is included
  if (!"PWGTP" %in% numeric_vars) {
    stop("PWGTP must be included in numeric_vars.")
  }

  # Check that at least one numeric variable other than PWGTP is included
  if (length(numeric_vars[numeric_vars != "PWGTP"]) < 1) {
    stop("At least one numeric variable other than PWGTP must be included.")
  }

  # Define valid categorical variables
  valid_categorical_vars <- c(
    "FER", "HHL", "SCH", "SCHL", "SEX"
  )

 # Check that at least one categorical variable was provided
  if (length(categorical_vars) < 1) {
    stop("At least one categorical variable must be included.")
  }

  # Check that all categorical variables are valid
  if (any(!categorical_vars %in% valid_categorical_vars)) {
    stop("One or more categorical variables are not valid.")
  }

  # Define valid geography levels
  valid_geographies <- c("State", "Division", "Region")

  # Check that exactly one valid geography level was provided
  if (
    length(geography) != 1 ||
    !geography %in% valid_geographies
  ) {
    stop("Geography must be State, Division, or Region.")
  }

  # Load the saved variable-value information
  geography_values <- readRDS(
    "data/census_variable_values.rds"
  )

  # Select the metadata for the requested year
  year_values <- geography_values[[as.character(year)]]

  # Determine the geography variable and a default value
  if (geography == "State") {

    api_geography <- "state"

    if (year %in% c(2021, 2022)) {
      metadata_geography <- "ST"
    } else {
      metadata_geography <- "STATE"
    }

    default_geography_value <- "37"

  } else if (geography == "Division") {

    api_geography <- "division"
    metadata_geography <- "DIVISION"
    default_geography_value <- "5"

  } else {

    api_geography <- "region"
    metadata_geography <- "REGION"
    default_geography_value <- "3"
  }

  # Use the default geography value when none is supplied
  if (is.null(geography_value)) {
    geography_value <- default_geography_value
  }

  # Check that only one geography value was provided
  if (length(geography_value) != 1) {
    stop("Only one geography value or 'all' may be specified.")
  }
  
  # Obtain the allowable geography codes from the saved metadata
  valid_geography_values <- names(
    year_values[[metadata_geography]]
  )

  # Allow the user to request all values for the geography
  if (tolower(as.character(geography_value)) == "all") {

    api_geography_value <- "*"

  } else {

    geography_value <- as.character(geography_value)

    # Add a leading zero to one-digit state codes
    if (
      geography == "State" &&
      nchar(geography_value) == 1
    ) {
      geography_value <- paste0("0", geography_value)
    }

    # Check that the requested geography value is valid
    if (!geography_value %in% valid_geography_values) {
      stop("The requested geography value is not valid.")
    }

    api_geography_value <- geography_value
  }

  # Combine the variables requested by the user
  variables <- unique(
    c(numeric_vars, categorical_vars)
  )

  # Convert the variable names into the format needed for the API
  variable_string <- paste(
    variables,
    collapse = ","
  )

  # Retrieve the Census API key
  census_key <- Sys.getenv("CENSUS_API_KEY")

  if (census_key == "") {
    stop("Census API key not found.")
  }

  # Build the API URL
  pums_url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums",
    "?get=", variable_string,
    "&for=", api_geography, ":", api_geography_value,
    "&key=", census_key
  )

  # Query the Census API
  response <- GET(pums_url)

  if (status_code(response) != 200) {
    stop("The Census API request was not successful.")
  }

  # Parse the returned JSON content
  pums_content <- content(
    response,
    as = "parsed",
    encoding = "UTF-8"
  )

  # Convert the returned content into a tibble
  pums_data <- parse_pums_response(pums_content)

  # Identify numeric variables that are not time variables
  regular_numeric_vars <- numeric_vars[
    !numeric_vars %in% c("JWAP", "JWDP")
  ]

  # Convert the regular numeric variables to numeric values
  for (variable in regular_numeric_vars) {
    pums_data[[variable]] <- as.numeric(
      pums_data[[variable]]
    )
  }

  # Identify the requested time variables
  time_vars <- numeric_vars[
    numeric_vars %in% c("JWAP", "JWDP")
  ]

  # Convert requested time variables to midpoint time values
  for (variable in time_vars) {

    # Obtain the Census time labels for the selected year
    time_labels <- unlist(
      year_values[[variable]]
    )
  
    # Match each Census code to its corresponding time range
    time_ranges <- unname(
      time_labels[pums_data[[variable]]]
    )
  
    # Convert each time range to its midpoint
    pums_data[[variable]] <- parse_time_midpoint(
      time_ranges
    )
  }
  
  # Convert the selected categorical variables to labeled factors
  for (variable in categorical_vars) {

    # Obtain the value labels for the selected survey year
    value_labels <- year_values[[variable]]

    # Convert the variable to a factor using the Census labels
    pums_data[[variable]] <- factor(
      pums_data[[variable]],
      levels = names(value_labels),
      labels = unlist(value_labels)
    )
  }
  
  # Add the census class to the returned tibble
  class(pums_data) <- c(
    "census",
    class(pums_data)
  )
  
  return(pums_data)
}

# Test the single-year function using its default values
pums_test <- get_pums_data()

# Examine the resulting tibble
head(pums_test)

# Examine the data types of the returned variables
str(pums_test)

# Test the function using the Region geography
pums_region_test <- get_pums_data(
  geography = "Region",
  geography_value = "3"
)

# Examine the resulting tibble
head(pums_region_test)
```

The single-year function successfully validates the requested year, variables, and geography and returns the selected PUMS data as a tibble. The default call returns data for North Carolina, while the second test confirms that the function can also return data for a specified Census region.

The returned PUMS data are then converted to appropriate data types. Standard numeric variables are converted from character values to numeric values, while categorical variables are converted to factors using the year-specific Census labels stored in the variable metadata. Time variables are excluded from the standard numeric conversion because they require separate processing.

```{r}
# Test the function with both Census time variables
pums_time_test <- get_pums_data(
  numeric_vars = c(
    "AGEP",
    "JWAP",
    "JWDP",
    "PWGTP"
  )
)

# Examine the first few observations
head(pums_time_test)

# Examine the data types and census class
str(pums_time_test)
class(pums_time_test)
```
```{r}
# Examine observations with available time values
head(
  pums_time_test[
    !is.na(pums_time_test$JWAP),
    c("JWAP", "JWDP")
  ]
)
```


