---
title: "ST558 Project 1"
author: "Aliyah Sheth"
format: html
editor: visual
---

## Introduction

This project develops functions for retrieving, processing, summarizing, and visualizing data from the Census Public Use Microdata Sample (PUMS) API. Data from the 2021 through 2024 surveys will be used. The project first processes Census variable metadata, then develops functions to query and combine PUMS data, followed by custom methods for weighted summaries and graphical displays.

## Part 1: Data Processing

### Parsing Variable Values

The first step is to obtain the variable descriptions for the Census PUMS data. The variable metadata are available as JSON files for each survey year. These files contain information about the variables and, where applicable, the values and labels associated with categorical variables.

The code below begins by querying the 2024 variable metadata to examine the 
structure of the returned JSON data. A status code of 200 indicates that the request was successful. 

```{r}
# Load package used to query the Census API
library(httr)

# URL for the 2024 PUMS variable metadata
url_2024 <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

# Query the Census API
response_2024 <- GET(url_2024)

# Check whether the request was successful
status_code(response_2024)
```

The returned JSON content is then parsed and examined to determine how the variable 
information is stored.

```{r}
# Parse the JSON response
metadata_2024 <- content(response_2024)

# Examine the top-level structure
names(metadata_2024)

# Examine the metadata for the SEX variable
metadata_2024$variables$SEX

# Examine the components stored for SEX
names(metadata_2024$variables$SEX)

# Variables required for the project
variables_2024 <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
  "STATE", "REGION", "DIVISION"
)

# Check whether any required variables are missing
variables_2024[
  !variables_2024 %in% names(metadata_2024$variables)
]
```
All variables required for the 2024 portion of the project are present in the Census metadata. 

#### Creating a Function to Parse Variable Values

After examining the structure of the 2024 variable metadata, a function is created to automate this process for the 2021 through 2024 survey years. The function identifies the appropriate state variable for each year, retrieves the corresponding JSON file, and extracts the value information for the variables required for this project.

```{r}
# Function to retrieve variable values for a specified survey year
get_variable_values <- function(year) {

  # Check that a valid survey year was provided
  if (!year %in% c(2021, 2022, 2023, 2024)) {
    stop("Year must be 2021, 2022, 2023, or 2024.")
  }

  # Determine the appropriate state variable name
  if (year %in% c(2021, 2022)) {
    state_variable <- "ST"
  } else {
    state_variable <- "STATE"
  }

  # Define the variables needed for the project
  variables <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_variable, "REGION", "DIVISION"
  )

  # Create the URL for the selected year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )

  # Query and parse the variable metadata
  response <- GET(url)

  if (status_code(response) != 200) {
    stop("The Census API request was not successful.")
  }

  metadata <- content(response)

  # Extract the available values for each required variable
  variable_values <- lapply(
    variables,
    function(variable) metadata$variables[[variable]]$values$item
  )

  # Assign variable names to the resulting list
  names(variable_values) <- variables

  return(variable_values)
}
```

The function is tested using the 2024 survey year, to ensure it produces the expected results. 

```{r}
# Test the function using the 2024 survey year
variable_values_2024 <- get_variable_values(2024)

# Examine the variables contained in the resulting list
names(variable_values_2024)

# Examine the values returned for SEX
variable_values_2024$SEX
```
#### Applying the Function to Parse Variable Values

The function is then applied to each of the four survey years to create a separate list of variable values for 2021, 2022, 2023, and 2024.

```{r}
# Create variable value lists for each survey year
variable_values_2021 <- get_variable_values(2021)
variable_values_2022 <- get_variable_values(2022)
variable_values_2023 <- get_variable_values(2023)
variable_values_2024 <- get_variable_values(2024)

# Combine the four yearly lists
variable_values <- list(
  "2021" = variable_values_2021,
  "2022" = variable_values_2022,
  "2023" = variable_values_2023,
  "2024" = variable_values_2024
)

# Examine the survey years stored in the list
names(variable_values)

# Check the state variable used for each survey year
names(variable_values$`2021`)
names(variable_values$`2024`)
```
The results confirm that the function correctly uses "ST" for the 2021 and 2022 surveys and "STATE" for the 2023 and 2024 surveys.

```{r}
# Save the variable value lists for later use
saveRDS(
  variable_values,
  file = "data/census_variable_values.rds"
)

# Read the saved file to verify that it was created successfully
variable_values_check <- readRDS(
  "data/census_variable_values.rds"
)

names(variable_values_check)
```
The saved object contains the variable-value information for each survey year and can be loaded later when categorical and geographic variables are converted to their appropriate labels.

### Parsing API Data

The next step is to query the PUMS API for person-level data. Before creating a general function, a single API request is examined to understand the structure of the data returned by the Census API.

```{r}
# Retrieve Census API key from the environment
census_key <- Sys.getenv("CENSUS_API_KEY")

# Check that an API key is available
if (census_key == "") {
  stop("Census API key not found.")
}

# Create a sample 2024 PUMS API request
pums_url <- paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  "?get=SEX,AGEP,PWGTP",
  "&for=state:37",
  "&key=", census_key
)

# Query the Census API
pums_response <- GET(pums_url)

# Check whether the request was successful
status_code(pums_response)
```
A status code of 200 indicates that the PUMS API request was successful. 

The returned content is then parsed to examine how the observations are organized. Although the project instructions indicate that an API key is not required, the Census Bureau began requiring API keys for all Census Data API queries in 2026. Therefore, an API key stored as an environment variable is used for the PUMS data queries in this project.

```{r}
# Parse the API response
pums_content <- content(
  pums_response,
  as = "parsed",
  encoding = "UTF-8"
)

# Examine the first few elements of the returned object
pums_content[1:3]

# Examine the first row returned by the API
pums_content[[1]]
```
#### Creating a Helper Function to Parse PUMS Data

The PUMS API returns the column names as the first element of the parsed response and the observations as the remaining elements. A helper function is created to convert this structure into a tibble with the appropriate column names.

```{r}
# Helper function to convert parsed PUMS content into a tibble
parse_pums_response <- function(pums_content) {

  # Extract the column names from the first element
  column_names <- unlist(pums_content[[1]])

  # Combine the remaining elements into rows
  data_matrix <- do.call(
    rbind,
    lapply(pums_content[-1], unlist)
  )

  # Convert to a data frame
  pums_data <- as.data.frame(
    data_matrix,
    stringsAsFactors = FALSE
  )

  # Assign the column names
  names(pums_data) <- column_names

  # Convert to a tibble
  pums_data <- tibble::as_tibble(pums_data)

  return(pums_data)
}

# Test the helper function
pums_tibble <- parse_pums_response(pums_content)

# Examine the resulting tibble
head(pums_tibble)
```
The helper function successfully converts the parsed API response into a tibble. At this stage, all returned variables are stored as character values; appropriate numeric, time, and categorical conversions will be performed in the subsequent data-processing function.
